---
title: "final-project-code"
author: "Parker Hicks"
date: "2023-11-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Notes
- The dataset only has 287 subject, not 316 as was stated in the Blood Storage Dataset Introduction
- It's unclear as to why not all patients are labeled as 1 for recurrence if the time to recurrence is recorded for all patients.

## Load libraries
```{r}
library(ggplot2)
library(ggmosaic)
library(dplyr)
library(DescTools)
library(coin)
```

## Read data
```{r}
bsd <- read.csv("./data/Blood_Storage.csv")
bsd <- na.omit(bsd)
```

## Mutate data for contingency table readability
```{r}
bsd <- bsd %>%
  mutate(bGS_risk_thresh2 = ifelse(bGS < 2, "low_risk", "high_risk")) %>%
  mutate(bGS_risk_thresh3 = ifelse(bGS <= 2, "low_risk", "high_risk")) %>%
  mutate(OrganPresence = ifelse(OrganConfined == 0, "No", "Yes")) %>%
  mutate(Recurrence_ = ifelse(Recurrence == 0, "No", "Yes"))
  
low_bGS <- subset(bsd, bGS_risk_thresh2 == "low_risk")
high_bGS <- subset(bsd, bGS_risk_thresh2 == "high_risk")
```

# Cohort summary
```{r}
bsd$Recurrence <- as.numeric(bsd$Recurrence)

# Means of variables
bsd %>% 
  group_by(bGS) %>%
  summarise(across(where(is.numeric), ~ mean(.x)))

# Sums of variables
bsd %>% 
  group_by(bGS) %>%
  summarise(across(where(is.numeric), sum))

# Sample size of groups
bsd %>% 
  group_by(bGS) %>%
  summarise(N = length(bGS))
```

## Plot contingency table
```{r}
ggplot(data=bsd) +
  theme_minimal() +
  geom_mosaic(aes(x=bGS_risk_thresh2, fill=Recurrence_)) +
  xlab("Biopsy Gelason Score (HR threshold = 2)") +
  ylab("Biochemical Recurrence") +
  theme(axis.text=element_text(size=11), 
        axis.title=element_text(size=13)
        )

# ggsave("imgs/mosaic_recurrence_bGS=2.png", dpi=500)
```

## Plot contingency table
```{r}
ggplot(data=bsd) +
  theme_minimal() +
  geom_mosaic(aes(x=bGS_risk_thresh3, fill=Recurrence_)) +
  xlab("Biopsy Gelason Score (HR threshold = 3)") +
  ylab("Biochemical Recurrence") +
  theme(axis.text=element_text(size=11), 
        axis.title=element_text(size=13)
        )

# ggsave("imgs/mosaic_recurrence_bGS=3.png", dpi=500)
```

## Define risk statistics functions
```{r}
# Risk Difference
calculateRiskDifference <- function(a, b, c, d) {
  p1 <- a / (a+b)
  p2 <- c / (c+d)
  
  return(p1 - p2)
}

riskDifferenceSE <- function(a, c, n1, n2) {
  p <- (a+c) / (n1+n2)
  se <- sqrt(p*(1-p)*((1/n1) + (1/n2)))
  
  return(se)
}

# Risk Ratio
calculateRiskRatio <- function(a, b, c, d) {
  p1 <- a / (a+b)
  p2 <- c / (c+d)
  
  return(p1 / p2)
}

riskRatioSE <- function(a, b, c, d) {
  n1 <- a+b
  n2 <- c+d
  
  se <- sqrt((b / (a*n1)) + (d / (c*n2)))
  return(se)
}

calculateOddsRatio <- function(a, b, c, d) {
  OR <- (a*d) / (b*c)
  return(OR)
}

oddsRatioSE <- function(a, b, c, d) {
  se <- sqrt((1 / a) + (1 / b) + (1 / c) + (1 / d))
  return(se)
}
```

## Calculate risk statistics
```{r}
risk_table <- function(outcome, exposure, data) {
  # Set contingency table parameters
  contingency_tbl <- table(data[[outcome]], data[[exposure]])
  a <- contingency_tbl["Yes", "high_risk"]
  b <- contingency_tbl["Yes", "low_risk"]
  c <- contingency_tbl["No", "high_risk"]
  d <- contingency_tbl["No", "low_risk"]

  # Set critical Z value
  Z <- 1.96
  
  # Calculate values
  risk_diff <- calculateRiskDifference(a, b, c, d)
  risk_diff_se <- riskDifferenceSE(a, c, a + b, c + d)
  risk_diff_ci <- c(risk_diff - (risk_diff_se * Z), 
                    risk_diff + (risk_diff_se * Z)
  )

  risk_ratio <- calculateRiskRatio(a, b, c, d)
  risk_ratio_se <- riskRatioSE(a, b, c, d)
  risk_ratio_ci <- c(exp(log(risk_ratio) - (Z * risk_ratio_se)), 
                     exp(log(risk_ratio) + (Z * risk_ratio_se))
  )

  odds_ratio <- calculateOddsRatio(a, b, c, d)
  odds_ratio_se <- oddsRatioSE(a, b, c, d)
  odds_ratio_ci <- c(exp(log(odds_ratio) - (Z * odds_ratio_se)), 
                     exp(log(odds_ratio) + (Z * odds_ratio_se))
  )

  # Create a data frame
  result_table <- data.frame(
    Measure = c("Risk Difference", "Risk Ratio", "Odds Ratio"),
    Value = c(risk_diff, risk_ratio, odds_ratio),
    Lower_CI = c(risk_diff_ci[1], risk_ratio_ci[1], odds_ratio_ci[1]),
    Upper_CI = c(risk_diff_ci[2], risk_ratio_ci[2], odds_ratio_ci[2])
  )
  
  return(result_table)
}

risk_table_bGS_thresh2 <- risk_table(outcome="Recurrence_", exposure="bGS_risk_thresh2", data=bsd)
risk_table_bGS_thresh3 <- risk_table(outcome="Recurrence_", exposure="bGS_risk_thresh3", data=bsd)

risk_table_bGS_thresh2
risk_table_bGS_thresh3
```

# Plot odds ratios
```{r}
df <- data.frame(yAxis = c("bGS HR=2", "bGS HR=3"),
  boxOdds = 
   c(5.8604651, 6.1488971),
  boxCILow = 
   c(2.8384892, 2.7990454),
  boxCIHigh = 
  c(12.0997647, 13.507796)
)


ggplot(df, aes(x = boxOdds, y = yAxis)) +
  geom_vline(aes(xintercept = 1), size = .75, linetype = 'dashed') +
  geom_errorbarh(aes(xmax = boxCIHigh, xmin = boxCILow), size = 1, height = 
      .2, color = 'gray50') +
  geom_point(size = 5, color = 'orange') +
  xlim(0, 15) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), axis.text=element_text(size=11), 
        axis.title=element_text(size=13),
        legend.position = "none") +
  ylab('bGS risk threshold') +
  xlab('Odds ratio')

# ggsave("imgs/OR_bGS_thresh.png", dpi=500)
```

## Compute permutation test
```{r}
bGS_thresh2 <- table(bsd$Recurrence_, bsd$bGS_risk_thresh2)
bGS_thresh2 <- bGS_thresh2[c(2, 1),]

bGS_thresh3 <- table(bsd$Recurrence_, bsd$bGS_risk_thresh3)
bGS_thresh3 <- bGS_thresh3[c(2, 1),]

fisher.test(bGS_thresh2)
fisher.test(bGS_thresh3)
```

## Plot Time to recurrence by bGS group
```{r}
ttr_subset <- subset(bsd, select=c("Recurrence", "bGS", "TimeToRecurrence"))
ttr_subset$Recurrence <- as.factor(ttr_subset$Recurrence)
ttr_subset$bGS <- as.factor(ttr_subset$bGS)
# ttr_subset <- filter(ttr_subset, Recurrence == 1)
ggplot(ttr_subset, aes(x=bGS, y=TimeToRecurrence, fill=bGS)) +
  geom_boxplot(size=.75) +
  geom_jitter(alpha=.5) +
  theme_minimal() +
  xlab("Biopsy Gleason score group") +
  ylab("Months to biochemical recurrence of PCa") +
  theme(axis.text=element_text(size=11), 
        axis.title=element_text(size=13),
        legend.position = "none")

# ggsave("imgs/bGS_ttr.png", dpi=500)
```

## Test for homogenetity of variance of time to recurrence
```{r}
bartlett.test(TimeToRecurrence ~ bGS, data=ttr_subset)
oneway.test(TimeToRecurrence ~ bGS, data=ttr_subset, var.equal = FALSE)
```

# Perform ANOVA and post-hoc LSD test
```{r}
bGS_aov <- aov(TimeToRecurrence ~ bGS, data=ttr_subset)
PostHocTest(bGS_aov, method=c("lsd"))
```

# Test normality of time to recurrence
```{r}
shapiro.test(bsd$TimeToRecurrence)
hist(bsd$TimeToRecurrence)
```

## Permutation test for the difference in time to recurrence from Gleason Scores
```{r}
set.seed(1234)
treatment <- as.factor(bsd$bGS_risk_thresh2)
outcome <- bsd$TimeToRecurrence

#Difference in means
original <- diff(tapply(outcome, treatment, mean))

results<-vector('list',10000)
for(i in 1:10000){
  x <- split(sample(bsd$TimeToRecurrence), rep(1:2, c(143,144)))
  results[[i]]<-mean(x[[1]]) - mean(x[[2]])  
}

permutation_df <- data.frame(perm_res = unlist(results))

# p-value
p <- (sum(unlist(results) > original) /10000) * 2
cat("Permutation test p-value:", p)

ggplot(permutation_df, aes(x = perm_res)) +
  geom_histogram(binwidth = 0.75, fill = "lightblue", color = "black", alpha=0.7) +
  labs(x = expression(bar("Time")[High~risk] - bar("Time")[Low~risk]), y = "Frequency") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text=element_text(size=11),
        axis.title=element_text(size=13)
        ) +
  xlim(-16, 16) +
  geom_vline(xintercept = original, color = "blue", linetype = "dashed") +
  annotate("text", x = original, y = 500, label = "Observed Mean Difference",  vjust = 0.5, color = "blue", size=4)

# ggsave("imgs/perm_meanDiff_recurrence_bGS=2.png", dpi=500)
```
# Compare p-value to independence test
```{r}
independence_test(outcome ~ treatment)
```

